\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-09-20 Manuscript/manuscript.tex
Reworked introduction, need to work on discussion still

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined reaction times for nine male athletes. The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds. Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Unfortunately, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
It is unknown how many athletes since the 2009 study could have potentially
benefited from this, but it is probable that there are others besides Allen who
could have. Additionally,
\citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been shown to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles. Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}. This 
contradicts the results of \citet{zhang2021correlation}, which indicated that 
year of competition has a significant effect on reaction times. Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article. A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}. Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}. While this may seem trivial, Allen was disqualified
only by 0.001 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men occurred 
between the ages of 26 and 29 \citep{tonnessen2013reaction}. This is notable 
because Allen was 27 years old at the 2022 World Track and Field Championships 
and likely at his physical peak.

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways. First we compare reaction times for athletes who 
have competed at additional competitions to the 2022 World Track and Field 
Championship. By comparing reaction times from the same athlete across multiple
competitions we are able to determine if the 2022 reaction times differed 
significantly from those of other competitions. Then we look at a longer time horizon
and model the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with two random effects: 
one effect for the year and a second effect for the heat nested within each year.
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

>>>>

The first objective of this paper is to determine whether there were 
abnormalities with athlete's reaction times at the 2022 World Track and Field
Championship.  Our analysis consists of quantitatively comparing reaction times for athletes who 
have competed at additional competitions to the 2022 World Track and Field 
Championship (Similar to a matched pairs design). By comparing reaction times from the same athlete across multiple
competitions we are able to determine if the 2022 reaction times differed 
significantly from those of other competitions.  This investigation, while
motivated by Allen's disqualifcation has wider impacts across track and field
that question if there is more variation than previously thought in track and
field timing equipment.

The second objective of this paper is to analyze whether the current
reaction time barrier of 0.1 seconds is fair, and in this regard there have been
many publications debating the human capacity for reaction times 
\of{This sounds funny but I'm not sure how to fix it}.
\citet{pain2007sprint} examined reaction times for nine male athletes. The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds. Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
These papers support increasing the reaction time barrier as they determined
that athletes are better than how IAAF thinks they are. Unfortunately, IAAF has
not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
Additionally, \citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been shown to be a reliable
measure of reaction time.

However, not all studies have been in favor of a decrease in the
reaction time barrier \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article. A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}. However it is
worth noting that \citet*{brosnan2017effects} and \citep{lipps2011implications}
were retroactive studies of athletes at events while \citet{pain2007sprint} and
\citet{komi2009iaaf} were experiments showing that there are various methods
of testing reaction time capability.

The second part of this paper will model the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with two random effects: 
one effect for the year and a second effect for the heat nested within each year.
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1).
